Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-13 paper/ms.tex
some changes

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must be written as a mixture of damped sinusoid functions.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)=(2\,\pi)^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

@@

In the case of $J$ damped sinusoids we can check for negative values of the PSD by solving for the
roots of the power spectrum, abbreviating with $z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
which may be solved exactly for the roots.

>>>>

In this paper, we present a class of GP models that enable likelihood
calculations that scale linearly with the number of data points
$\mathcal{O}(N)$ for one dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic.
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, robust, simple,
and fast.

@@

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we will refer to the argument within the sum as a ``\celerite\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)=(2\,\pi)^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

@@

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum, abbreviating with
$z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
which may be solved exactly for the roots.
